# Practical Machine Learning: Project

Welcome to my write up of the Practical Machine Learning Project. This time we would like to build a algorithm to predict the manner in which participants did excercises which thy tracked with their Jawbone Up, Nike FuelBand, Fitbit or other divices. While building the algorithm I will describe how I build the altorithm, how I use cross validation, what I think the expected out of sample error is and why I made the choices I made.

## Data Preparation
Let's start with the data. We will download and prepare it a little bit. The data is provided devided in training and test data sets.

```{r}
require(data.table)
setInternet2(TRUE)
url <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
trainingSet <- fread(url)
url <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"
testSet <- fread(url)
```

In another step we will exclude any variables from the model and training dataset which do have NA values.
```{r}
coloumsWhereNAsAre <- sapply(testSet, function (x) any(is.na(x) | x == ""))
coloumsWhereNoNAsAre <- !coloumsWhereNAsAre & grepl("belt|[^(fore)]arm|dumbbell|forearm", names(coloumsWhereNAsAre))
predCandidates <- names(coloumsWhereNAsAre)[coloumsWhereNoNAsAre]
#predCandidates

predictors <- c("classe", predCandidates)
trainingSet <- trainingSet[, predictors, with=FALSE]
#dim(trainingSet)
#names(trainingSet)
```

In a next step we need to make the classes variable a factor variable and we will split the training dataset again into a training and test dataset.

```{r}
trainingSet <- trainingSet[, classe := factor(trainingSet[, classe])]
#trainingSet[, .N, classe]

require(caret)
set.seed(1337)
toIncludeInTrainingSet <- createDataPartition(trainingSet$classe, p=0.6)
trainingSetTrain <- trainingSet[toIncludeInTrainingSet[[1]]]
trainingSetTest <- trainingSet[-toIncludeInTrainingSet[[1]]]
```

We will center and scale the left predictors in the training dataset, apply the centering and scaling to the test dataset and check vor variables with zero variance.

```{r}
X <- trainingSetTrain[, predCandidates, with=FALSE]
preProc <- preProcess(X)
preProc
XCS <- predict(preProc, X)
trainingSetTrainCS <- data.table(data.frame(classe = trainingSetTrain[, classe], XCS))

X <- trainingSetTest[, predCandidates, with=FALSE]
XCS <- predict(preProc, X)
trainingSetTestCS <- data.table(data.frame(classe = trainingSetTest[, classe], XCS))

nzv <- nearZeroVar(trainingSetTrainCS, saveMetrics=TRUE)
if (any(nzv$nzv)) nzv else message("No variables with near zero variance")
```

## Model building

We will use RandomForrest and parallel processing to find the best fitting predictors in a decent amount of time. 

```{r}
require(parallel)
require(doParallel)
cl <- makeCluster(detectCores() - 1)
registerDoParallel(cl)

require(randomForest)
require(e1071)
ctrl <- trainControl(classProbs=TRUE,
                     savePredictions=TRUE,
                     allowParallel=TRUE)

method <- "rf"
system.time(trainingModel <- train(classe ~ ., data=trainingSetTrainCS, method=method))

stopCluster(cl)
```

## Model Evaluation

To evaluate the model we will implement the model on the training test set to get the out of sample error. An error rate of about 3-4% or below would be pretty good.

```{r}
prediction <- predict(trainingModel, trainingSetTrainCS)
confusionMatrix(prediction, trainingSetTrainCS[, classe])

prediction <- predict(trainingModel, trainingSetTestCS)
confusionMatrix(prediction, trainingSetTestCS[, classe])

varImp(trainingModel)
trainingModel$finalModel

save(trainingModel, file="trainingModel.RData")
```

## Test Data Assesment

In the final step we will predict the classes of the 20 test datasets.

```{r}
load(file="trainingModel.RData", verbose=TRUE)

testSetCS <- predict(preProc, testSet[, predCandidates, with=FALSE])
prediction <- predict(trainingModel, testSetCS)
testSet <- cbind(prediction , testSet)
subset(testSet, select=names(testSet)[grep("belt|[^(fore)]arm|dumbbell|forearm", names(testSet), invert=TRUE)])
```

And finally the submission to Coursera.

```{r}
pml_write_files = function(x){
  n = length(x)
  path <- "P:/"
  for(i in 1:n){
    filename = paste0("problem_id_",i,".txt")
    write.table(x[i],file=file.path(path, filename),quote=FALSE,row.names=FALSE,col.names=FALSE)
  }
}
pml_write_files(lapply(prediction, as.character))
```

Thanks a lot for reading.

All the best,
Lennart